msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-12 paper.tex
Draft of arg stuff.

The \msprime~simulator generates tree sequences by simulating realisations of the
Ancestral Recombination Graph (ARG).
By default, only the marginal genealogy at each site is recorded in the tree sequence
output.
Marginal genealogies are sufficient for e.g.~simulating and storing sequence data,
and involve fewer edges than nodes than storing a realisation of the full ARG.
However, the mapping from ARG realisations to marginal tree sequences is not lossless.
For example, times of recombination events can be recovered from the former
but not the latter.
The same is true of the number of extant ancestors at each time in the past.
In addition to the default behaviour, we support the retention of the full
ARG realisation using the \texttt{record\_full\_arg} option.
This option augments the default tree sequence data structure with three
new kinds of nodes:
\begin{itemize}
\item Each recombination event is marked by two nodes with identical birth times
given by the time of the recombination event, and marked with the
\texttt{NODE\_IS\_RE\_EVENT} flag.
Each node represents an ancestor of the recombinant child node, and edges connect
the relevant segments of genome from the child node to these two ancestors.
\item We store common ancestor nodes even when the corresponding merger event
does not result in coalescence of ancestral material.
The resulting nodes are marked with the \texttt{NODE\_IS\_CA\_EVENT} flag.
\item When multiple populations are simulated, migration events are depicted by nodes
with the \texttt{NODE\_IS\_MIG\_EVENT} flag, and with birth times equal to the time of
the migration.
The population flag of the migration node identifies the population to which the lineage
belongs after the reverse time migration event.
\end{itemize}

The likelihood, i.e.~the joint sampling probability of the ancestry and mutations,
is an example of a function which is intractable when the ancestry is stored as a
marginal tree sequence, but tractable when the ancestry is given by a full
ARG realisation.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as Metropolis-Hastings
\citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.

We provide two functions to facilitate likelihood-based inference.
Both are implemented only for the simplest case of the standard ARG with a
constant population size, and require tree sequences compatible with the
\texttt{record\_full\_arg} option as their arguments.
\begin{itemize}
\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
per-generation recombination probability \texttt{r} and population size \texttt{N}.
Specifically, the function returns the logarithm of
\begin{equation*}
\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
		+ r l_i \Big] t_i  \Bigg),
\end{equation*}
where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
in that interval that would split ancestral material should they recombine,
$q$ is the total number of events in the tree sequence $\texttt{ts}$,
$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
of time intervals which end in a recombination, and $g_i$ is the corresponding
\emph{gap}:  the length of material in which the recombination at the end of the
$i$th period could have taken place without affecting the resulting genealogy.
A recombination in ancestral material will always have $g_i = 1$, while a recombination
in trapped non-ancestral material can result in other values of $g_i$.
For a continuous model of the genome and a recombination in ancestral material,
we set $g_i = 1$ and interpret the result as a density.
\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
natural logarithm of the probability of the mutations recorded in the tree sequence
\texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
a normalising constant which depends on the pattern of mutations,
but not on the tree sequence or the per-site, per-generation mutation
probability \texttt{m}.
Specifically, the function returns the logarithm of
\begin{equation*}
e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
\prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
\end{equation*}
where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
total branch length on which $\gamma$ could have arisen while appearing on all
of the leaves of \texttt{ts} it does, and on no others.
Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
edges.
\end{itemize}

\begin{comment}
Previous draft of the ARG feature retained here as a comment.

The definition of ARGs is vague and confusing.
\cite{minichiello2006mapping} define it as the data structure
rather than the original stochastic
process~\citep{griffiths1991two,griffiths1997ancestral}.
\cite{rasmussen2014genome} have a slightly different approach.
Sometimes the ARG is useful.
The tree sequence output for \msprime\ can be turned into
an ARG using the \texttt{record\_full\_arg} option, where we store
recombination nodes.

We also provide an interface to compute the likelihood of a given ARG
under the~\cite{kuhner2000maximum} model.
\end{comment}

>>>>

The term Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a graphical
representation of the coalescent with recombination stochastic process. This
formulation was complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of the recombination process. In Griffiths' ARG formulation, we
regard a realisation of the coalescent with recombination as a graph in which
nodes represent common ancestor or recombination events, and edges represent
lineages through which ancestral material flows. We may consider the ``big''
ARG, in which we wait an exponentially large time until all sampled lineages
coalescence into a single ancestor~\citep{ethier1990two}, or the ``little'' ARG
in which we wait until all marginal trees have fully coalescenced (like
Hudson's algorithm).

% Seems somewhat unfair to single out Iain and Alywyn here, but a systematic
% review would be long and tedious, and I'm sure they won't be too upset about it.
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean a record of genetic ancestry as a graph; for example,
\cite{mathieson2020ancestry} informally define ARGs as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unhelpful, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little
to clarify. As we shall see, there are situations in which the precise form of the
graph is important, and conflating high-level representation of genetic
ancestry with this precise mathematical object is not helpful.
One may wonder if a more generic term such as ``genetic pedigree''
would be a more useful description, in such cases where the precise form
of the encoding is not important.

% ~\cite[.e.g.][]{minichiello2007mapping}
% \cite{minichiello2006mapping} are somewhat more explicit
% "An ARG provides a record of all coalescence and recombination events since the
% divergence of the sequences under study and specifies a complete genealogy at
% each genomic position
% \citep{rasmussen2014genome}
For our purposes here, an ARG is a realisation of the coalescent with
recombination, in the Griffiths sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
essentially works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see section XX),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.
This approach is one of the key reasons for \msprime's efficiency.

Although the tree sequences output by \msprime\ contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occured, information
is lost in this mapping down from ARG space. In particular, we lose
information about individuals who were common ancestors but where
no coalescences occured and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the compete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) in the tree sequence.
Fig.~\ref{fig-recombination-perf}B shows that these extra ARG nodes
and edges quickly come to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.

One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
(encoded using the representation of the \texttt{record\_full\_arg} option)
and mutational patterns under
the standard coalescent and infinite sites mutation model.

% We provide two functions to facilitate likelihood-based inference.
% Both are implemented only for the simplest case of the standard ARG with a
% constant population size, and require tree sequences compatible with the
% \texttt{record\_full\_arg} option as their arguments.
% \begin{itemize}
% \item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of
% the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
% per-generation recombination probability \texttt{r} and population size \texttt{N}.
% Specifically, the function returns the logarithm of
% \begin{equation*}
% \Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
% 	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 }
% 		+ r l_i \Big] t_i  \Bigg),
% \end{equation*}
% where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
% $k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
% in that interval that would split ancestral material should they recombine,
% $q$ is the total number of events in the tree sequence $\texttt{ts}$,
% $q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices
% of time intervals which end in a recombination, and $g_i$ is the corresponding
% \emph{gap}:  the length of material in which the recombination at the end of the
% $i$th period could have taken place without affecting the resulting genealogy.
% A recombination in ancestral material will always have $g_i = 1$, while a recombination
% in trapped non-ancestral material can result in other values of $g_i$.
% For a continuous model of the genome and a recombination in ancestral material,
% we set $g_i = 1$ and interpret the result as a density.
% \item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
% natural logarithm of the probability of the mutations recorded in the tree sequence
% \texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
% a normalising constant which depends on the pattern of mutations,
% but not on the tree sequence or the per-site, per-generation mutation
% probability \texttt{m}.
% Specifically, the function returns the logarithm of
% \begin{equation*}
% e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
% \prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
% \end{equation*}
% where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
% in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
% total branch length on which $\gamma$ could have arisen while appearing on all
% of the leaves of \texttt{ts} it does, and on no others.
% Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
% mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
% edges.
% \end{itemize}

% \begin{comment}
% Previous draft of the ARG feature retained here as a comment.

% The definition of ARGs is vague and confusing.
% \cite{minichiello2006mapping} define it as the data structure
% rather than the original stochastic
% process~\citep{griffiths1991two,griffiths1997ancestral}.
% \cite{rasmussen2014genome} have a slightly different approach.
% Sometimes the ARG is useful.
% The tree sequence output for \msprime\ can be turned into
% an ARG using the \texttt{record\_full\_arg} option, where we store
% recombination nodes.

% We also provide an interface to compute the likelihood of a given ARG
% under the~\cite{kuhner2000maximum} model.
% \end{comment}
